Natural Selection in the 20th Century

### 2020-10-13 why-negative-selection.Rmd
New intro; age-controlled n_partners regression

---
title: "Negative Selection in the 20th Century"
author: "David Hugh-Jones, Abdel Abdellaoui"
date: "July 2020"
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
---

@@

Existing work suggests that natural selection is taking place in modern populations,
but that effect sizes are small. In this paper we investigate natural selection
in the UK over the twentieth century, using data from UK Biobank. We use
questions on number of children and number of siblings to reconstruct two
generations of family size per respondent. We then relate these to 33 different
polygenic scores for a variety of  characteristics.

Natural selection is taking place on most polygenic scores, and is concentrated
among population subgroups that are under-represented in the sample, including
less-educated people and poorer households. When we weight the sample to correct
for ascertainment bias, many effect sizes are considerably larger.

Among mothers who were older at first birth, and among parents with fewer sexual
partners, effect signs are reversed. The effects of PGS can be decomposed into
two channels: an effect on age at first live birth, and an opposite-signed
effect on number of children conditional on age at first live birth. This
pattern can be explained by economic theories of fertility. In these, higher
potential earnings have two opposite effects on fertility: a
fertility-increasing income effect (higher income makes children more
affordable), and a fertility-lowering substitution effect (time spent on
childrearing has a higher cost in foregone earnings). Among couples who can take
advantage of coordination in childcare activities, the income effect dominates;
among single parents, or those in unstable relationships, the substitution effect
dominates.

@@

These estimates may suffer from ascertainment bias within the UK Biobank sample.
Respondents are higher income and better educated than the UK population, and
they may also differ on other unobserved characteristics.

To examine the effects of sampling, we regress respondents' polygenic
scores on their number of children:

@@

The effect size $\beta$ measures the strength of natural selection within the
sample (or a given subgroup). In fact, since polygenic scores are normalized to
mean 0, $\beta$ is the expected polygenic score among children of the sample.

Figures \@ref(fig:plot-educ-level) and \@ref(fig:plot-income) plot effect sizes
for each polygenic score among different subgroups of the sample. Figure
\@ref(fig:plot-educ-level) groups respondents by age of completing full-time
education. Figure \@ref(fig:plot-income) groups respondents by household income.
Effect sizes are larger and more significant for the lowest income category, and
for the lowest education category. Note that the overall effect is not a simple
average of the effect among the different subgroups, because polygenic scores 
may also shift respondents between the subgroups. For example, a high 
PGS for educational attainment may increase the age at which a respondent 
leaves school.

>>>>

---
title: "Natural Selection in the 20th Century"
author: "David Hugh-Jones, Abdel Abdellaoui"
date: "July 2020"
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
---

@@

Geneticists have long hypothesized that natural selection is taking place in
modern human populations. Recent work confirms this using genome-wide analysis
[@Barban_2016; @beauchamp2016genetic; @kong2017selection; @sanjak2018evidence].
In particular, genetic variants associated with higher educational attainment
are being selected against, although effect sizes appear small.

Several gaps in our knowledge remain. First, while existing work examines
contemporary populations, we would like to extend the analysis over time, since
natural selection in the same direction over multiple generations
could lead to significant effects in the long run. Second, there are concerns
about ascertainment bias in the samples used. This could potentially bias
estimates of the magnitude of natural selection. Lastly, we know little about
the *mechanisms* behind natural selection in contemporary populations.

Here we attempt to answer these questions, using data from UK Biobank. We use
questions on respondents' number of children, and their number of siblings, to 
reconstruct two generations of family size. We then relate these to 33 different
polygenic scores for a variety of characteristics. This analysis provides a 
broader picture of natural selection in the UK than has previously been available.

Our key results are as follows. First, comparing natural selection across the
two generations, there are few signs of change in size or direction over time.
Most polygenic scores are selected in the same direction in both generations. We
also see little evidence of changes across birth cohorts within either
generation.

Second, natural selection is concentrated among population subgroups which are
under-represented in the sample, including less-educated people and poorer
households. When we weight the sample to correct for ascertainment bias, many
effect sizes are larger, sometimes much larger. We emphasize that our weights
are only a partial solution to ascertainment bias in the sample. As a result,
much uncertainty remains about the strength of natural selection in the UK
population, but we can be reasonably sure that it is larger than naïve estimates
suggest.

Lastly, patterns in natural selection give us clues to the underlying
mechanism. In particular, among mothers who were older at their first birth, and
among parents with fewer sexual partners, the direction of natural selection is
*reversed*: for example, among these subgroups, higher polygenic scores (PGS)
for educational attainment are associated with having more offspring, rather
than fewer. The effects of PGS can be decomposed into two channels: an effect on
age at first live birth, and an opposite-signed effect on number of children
conditional on age at first live birth.

This pattern can be explained by economic theories of fertility
[@becker1960economic]. In these, higher potential earnings have two opposite
effects on fertility: a fertility-increasing *income effect* (higher income
makes children more affordable), and a fertility-lowering *substitution effect*
(time spent on childrearing has a higher cost in foregone earnings). Among
couples who can take advantage of coordination in childcare activities, the
income effect dominates; among single parents, or those in unstable
relationships, the substitution effect dominates. Supporting this explanation,
polygenic scores' raw effects on fertility are negatively correlated with their
effects on income; on the other hand, controlling for age at first live birth,
scores' effects on fertility and on household income are *positively*
correlated. These two results are compatible with the substitution effect and the
income effect respectively.

@@

These estimates may suffer from ascertainment bias within the UK Biobank sample.
Respondents are higher income and better educated than the UK population, and
they may also differ on other unobserved characteristics [@10.1093/aje/kwx246].

To examine the effects of sampling, we start by measuring the strength of
natural selection within different subgroups. To do this we regress respondents' 
polygenic scores on their number of children:

@@

The effect size $\beta$ measures the strength of natural selection within the
sample (or a given subgroup). In fact, since polygenic scores are normalized to
mean 0, $\beta$ is the expected polygenic score among children of the sample or
subgroup.

Figures \@ref(fig:plot-educ-level) and \@ref(fig:plot-income) plot effect sizes
for each polygenic score. Figure \@ref(fig:plot-educ-level) groups respondents
by age of completing full-time education. Figure \@ref(fig:plot-income) groups
respondents by household income. Effect sizes are larger and more significant
for the lowest income category, and for the lowest education category. Note that
the overall effect is not a simple average of the effect among the different
subgroups, because polygenic scores may also shift respondents between the
subgroups. For example, a high PGS for educational attainment may increase the
age at which a respondent leaves school.
